---
title: 'Learning Balto-Finnic dialects'
author: "Kaili Vesik"
date: '2023-11-21'
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
# don't show this output
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

We begin by setting options (output not shown) and loading the appropriate libraries.


```{r load}
# for data wrangling
library(tidyverse)
# for plotting
library(ggplot2)
library(cowplot)
# library(ggbeeswarm)
# library(vipor)
# library(viridis)
# library(ggstance)
# library(ggrepel)
library(scales)
```

## Constraint set

Create the vectors containing the faithfulness and (many) markedness constraint names.

```{r constraintnames}
identcons <- c("IdBkSyl1", "IdBk")
maxcons <- c("MaxSyl1", "Max")
nonempty <- c("NonEmpty")

frontsets <- c("F1", "F3", "F4", "F5")
backsets <- c("B1", "B2", "B3", "B5")
starfronts <- paste("star", frontsets, sep="")
starbacks <- paste("star", backsets, sep="")

starfrontbacks <- c()
starbackfronts <- c()
for (fr in frontsets) {
  for (bk in backsets) {
    starfrontbacks <- c(starfrontbacks, paste(fr, bk, sep=""), paste(fr, bk, sep="_"))
    starbackfronts <- c(starbackfronts, paste(bk, fr, sep=""), paste(bk, fr, sep="_"))
  }
}
starfrontbacks <- paste("star", starfrontbacks, sep="")
starbackfronts <- paste("star", starbackfronts, sep="")
```


## Column titles

Generate options for column title vectors, according to experiment type.

```{r coltitles}

num_out_in <- c("trialnum", "generated", "heard")

identconswchanges <- paste0(rep(identcons, each=2), c("_change", ""))
maxconswchanges <- paste0(rep(maxcons, each=2), c("_change", ""))
nonemptywchanges <- paste0(rep(nonempty, each=2), c("_change", ""))
starfrontswchanges <- paste0(rep(starfronts, each=2), c("_change", ""))
starbackswchanges <- paste0(rep(starbacks, each=2), c("_change", ""))
starfrontbackswchanges <- paste0(rep(starfrontbacks, each=2), c("_change", ""))
starbackfrontswchanges <- paste0(rep(starbackfronts, each=2), c("_change", ""))

# append different combinations of column titles for different types of simulations
coltitlesall_max0nonempty0 <- c(num_out_in, identconswchanges, starfrontswchanges, starbackswchanges, starfrontbackswchanges, starbackfrontswchanges)
coltitlesall_max1nonempty0 <- c(num_out_in, identconswchanges, maxconswchanges, starfrontswchanges, starbackswchanges, starfrontbackswchanges, starbackfrontswchanges)
coltitlesall_max1nonempty1 <- c(num_out_in, identconswchanges, maxconswchanges, nonemptywchanges, starfrontswchanges, starbackswchanges, starfrontbackswchanges, starbackfrontswchanges)

# select different combinations of columns for different types of simulations
coltitlestoselect_max0nonempty0 <- c("trialnum", identcons, starfronts, starbacks, starfrontbacks, starbackfronts)
coltitlestoselect_max1nonempty0 <- c("trialnum", identcons, maxcons, starfronts, starbacks, starfrontbacks, starbackfronts)
coltitlestoselect_max1nonempty1 <- c("trialnum", identcons, maxcons, nonempty, starfronts, starbacks, starfrontbacks, starbackfronts)

```


## Data

Load the data from file, and omit columns irrelevant to this analysis.

```{r data}
prepare_data <- function(filename) {
  if (!grepl("wdel", filename)) {
    coltitlesall <- coltitlesall_max0nonempty0
    coltitlestoselect <- coltitlestoselect_max0nonempty0
  } else if (grepl("-ne", filename)) {
    coltitlesall <- coltitlesall_max1nonempty1
    coltitlestoselect <- coltitlestoselect_max1nonempty1
  } else {
    coltitlesall <- coltitlesall_max1nonempty0
    coltitlestoselect <- coltitlestoselect_max1nonempty0
  }
  
  # load data from file, name columns, remove unwanted columns
  df <- read_tsv(paste0("..\\simulation_outputs\\",filename), 
                 # https://stackoverflow.com/questions/20641222/reading-csv-using-r-where-header-is-on-the-second-line
                 skip = 1,
                 # https://readr.tidyverse.org/reference/read_delim.html
                 col_names = coltitlesall,
                 col_types = list(
                   trialnum = col_integer(),
                   generated = col_character(),
                   heard = col_character(),
                   .default = col_double()))

  df <- 
    df %>%
    select(all_of(coltitlestoselect))
  
  # fill in missing data and adjust trial number for initial state
  df$trialnum[1] <- 0  # initial values
  df <- 
    df %>%
    fill(all_of(coltitlestoselect)) %>%
    fill(all_of(coltitlestoselect), .direction = "up")
  
  
  # just keep the last entry for each trial number 
  # (so, after both the learning adjustment and a possible a priori ranking adjustment)
  df <-
    df %>%
    arrange(trialnum) %>%
    group_by(trialnum) %>% 
    summarise_all(last) %>%
    ungroup() %>%
    arrange(trialnum)
  
  # pivot from wide to long, so we don't have a separate column for each constraint
  df <-
    df %>% 
    pivot_longer(!trialnum, names_to="constraint", values_to="value")
  
  # grab the last value (ie, the one in the highest trialnum) of each constraint
  finalvalues <- df %>% 
    group_by(constraint) %>% 
    top_n(1, trialnum) %>%
    pull(value) %>%
    round(digits=2)
  
  constraint_levels <- coltitlestoselect[-1]  # not including trialnum
  constraint_labels <- constraint_levels # TODO for now...
  constraint_labels <- str_replace(constraint_labels, "star", "*")
  constraint_labels <- str_replace(constraint_labels, "_", "...")
  constraint_labels <- str_replace(constraint_labels, "Bk", "(Bk)")
  # constraint_labels <- c("*F", "*ɤ", "*e", "Id-Syl1(Bk)", "Id(Bk)", "Agr(Bk)", "GMH(F)", "GMH(ɤ)", "GMH(e)")
  
  constraint_labelswithvalues <- paste(constraint_labels, "→", finalvalues)
  df$constraint <- factor(df$constraint, levels=constraint_levels, labels=constraint_labelswithvalues)
   
  return(df)
}
```
## Plots

Prepare and plot the trajectories. 

```{r plots}

removeconvalue <- function(labels) {
  cons <- rep("", each=length(labels))
  for (i in 1:length(labels)) {
    cons[i] <- strsplit(labels[i], split = " ")[[1]][1]
  }
  return(cons)
}

map_colours <- function(lang, cons, labels) {
  # set up the constraint / colour map in its default form (common across languages)
  colourmap <- rep("darkgrey", each=length(cons))
  names(colourmap) <- cons
  colourmap["Id(Bk)Syl1"] <- "cyan"
  if ("Max" %in% names(colourmap)) {
    colourmap[c("Max", "MaxSyl1")] <- "cyan"
  }
  colourmap["Id(Bk)"] <- "yellow"
  
  # per-language colour codings
  if (tolower(lang) == 'ne') {
    # full bans
    colourmap[c("*B1")] <- "cyan"
    # vowel harmony
    # n/a 
    # positional restrictions
    colourmap[c("*B2", "*F3")] <- "green"
    # must be inactive (outranked by Id(Bk))
    colourmap[c("*B3", "*B5", "*F4", "*F5")] <- "orange"
    colourmap[c("*F1B3", "*F1...B3", "*F1B5", "*F1...B5", "*F3B3", "*F3...B3", "*F3B5", "*F3...B5", "*F4B5", "*F4...B5", "*F5B3", "*F5...B3", "*F5B5", "*F5...B5", "*B2F4", "*B2...F4", "*B2F5", "*B2...F5", "*B3F4", "*B3...F4", "*B3F5", "*B3...F5", "*B5F4", "*B5...F4", "*B5F5", "*B5...F5")] <- "orange"
  } else if (tolower(lang) == 'fi') {
    # full bans
    colourmap[c("*B2")] <- "cyan"
    # vowel harmony
    colourmap[c("*F1B5", "*F1...B5", "*F3B5", "*F3...B5", "*B3F3", "*B3...F3", "*B5F3", "*B5...F3")] <- "cyan"
    # positional restrictions
    # n/a
    # must be inactive (outranked by Id(Bk))
    colourmap[c("*B3", "*B5", "*F1", "*F3", "*F4", "*F5")] <- "orange"
    colourmap[c("*F4B3", "*F4...B3", "*F4B5", "*F4...B5", "*F5B3", "*F5...B3", "*F5B5", "*F5...B5", "*B3F4", "*B3...F4", "*B3F5", "*B3...F5", "*B5F4", "*B5...F4", "*B5F5", "*B5...F5")] <- "orange"
  } else if (tolower(lang) == 'ns') {
    # full bans
    # n/a
    # vowel harmony
    colourmap[c("*F1B5", "*F1...B5", "*F3B5", "*F3...B5", "*F4B5", "*F4...B5", "*B1F4", "*B1...F4", "*B2F4", "*B2...F4", "*B3F4", "*B3...F4", "*B5F4", "*B5...F4")] <- "cyan"
    # positional restrictions
    colourmap[c("*B1")] <- "green"
    # must be inactive (outranked by Id(Bk))
    colourmap[c("*B2", "*B3", "*B5", "*F1", "*F3", "*F4", "*F5")] <- "orange"
    colourmap[c("*F5B2", "*F5...B2", "*F5B3", "*F5...B3", "*F5B5", "*F5...B5", "*B2F5", "*B2...F5", "*B3F5", "*B3...F5", "*B5F5", "*B5...F5")] <- "orange"
  } else if (tolower(lang) == 'ss') {
    # full bans
    # n/a
    # vowel harmony
    colourmap[c("*F1B2", "*F3B2", "*B1F1", "*B1...F1", "*B1F3", "*B1...F3", "*B1F4", "*B1...F4", "*B2F1", "*B2...F1", "*B2F3", "*B2...F3", "*B2F4", "*B2...F4", "*B3F1", "*B3...F1", "*B3F3", "*B3...F3", "*B3F4", "*B3...F4", "*B5F1", "*B5...F1", "*B5F3", "*B5...F3", "*B5F4", "*B5...F4")] <- "cyan"
    # positional restrictions
    colourmap[c("*B1", "*F1")] <- "green"
    # must be inactive (outranked by Id(Bk))
    colourmap[c("*B2", "*B3", "*B5", "*F1", "*F3", "*F4", "*F5")] <- "orange"
    colourmap[c("*F5B2", "*F5...B2", "*F5B3", "*F5...B3", "*F5B5", "*F5...B5", "*B2F5", "*B2...F5", "*B3F5", "*B3...F5", "*B5F5", "*B5...F5")] <- "orange"
  }
  # get the arrows and the ending values in there, since that's the format that ggplot will recognize
  names(colourmap) <- labels
  return(colourmap)
}

plot_trajectories <- function(df, colourmap, lang, trialsperbatch, experimentnum) {

  trajectoryplot <-
    ggplot(data=df, mapping=aes(x=trialnum, y=value, group=constraint, colour=constraint)) +
    geom_line(linewidth=0.6, show.legend=FALSE) +
    # geom_line(aes(color=fct_reorder2(constraint, trialnum, value, .fun = last2)), linewidth=0.6, show.legend=FALSE) +  # can replace all the fct business with just "constraint"
    ggtitle(paste0(lang, " learning trajectories over 4x", trialsperbatch, " trials (exp ", lang, experimentnum, ")")) +
    theme(axis.text=element_text(size=11, colour="black"),
        plot.title = element_text(hjust = 0.5),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank()) +
    labs(y= "Constraint value", x = "Learning trial #") +
    # https://tidyverse.github.io/ggplot2-docs/reference/scale_manual.html
    scale_colour_manual(values=colourmap, aesthetics="colour", na.value="red") +
    # scale_color_viridis_d(name = "constraint") +
    scale_x_continuous(trans="log10", labels=comma_format(accuracy=1)) +
    # TODO maybe specify limits dependent on actual range of ending values?
    scale_y_continuous(limits=c(50,150), breaks=seq(from=50, to=150, by=10))
    # scale_y_continuous(sec.axis = sec_axis(~ ., breaks = data_ends2)) +
  # geom_text_repel(
  #   aes(label = constraint), data = data_ends, 
  #   size = 3
  #   )
    
  print(trajectoryplot)
}

plot_legends <- function(df, colourmap, lang, trialsperbatch, experimentnum) {

  legendplot <-
    ggplot(data=df, mapping=aes(x=trialnum, y=value, group=constraint, colour=constraint)) +
    # https://stackoverflow.com/questions/54536094/ggplot-order-legend-using-last-values-on-x-axis
    geom_line(aes(color=fct_reorder2(constraint, trialnum, value, .fun = last2)), linewidth=0.6, show.legend=TRUE) +
    # ggplot(data=df, mapping=aes(x=trialnum, y=value)) +
    # geom_line(aes(color=colourmap), linewidth=0.6, show.legend=TRUE) + 
    # geom_line(aes(color=fct_reorder2(constraint, trialnum, value, .fun = last2)), linewidth=0.6, show.legend=TRUE) +  # can replace all the fct business with just "constraint"
    ggtitle(paste0(lang, experimentnum)) +
    theme(axis.text=element_text(size=11, colour="black"),
        plot.title = element_text(hjust = 0.5),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank()) +
    labs(y= "Constraint value", x = "Learning trial #") +
    scale_colour_manual(values=colourmap, aesthetics="colour", na.value="red") +
    # scale_color_viridis_d(name = "constraint") +
    scale_x_continuous(trans="log10", labels=comma_format(accuracy=1)) +
    scale_y_continuous(limits=c(50,150), breaks=seq(from=50, to=150, by=10))
    # scale_y_continuous(sec.axis = sec_axis(~ ., breaks = data_ends2)) +
  # geom_text_repel(
  #   aes(label = constraint), data = data_ends, 
  #   size = 3
  #   )
  # https://stackoverflow.com/questions/12041042/how-to-plot-just-the-legends-in-ggplot2
  leg <- get_legend(legendplot)
  legendonly <- ggdraw(plot_grid(NULL, leg, ncol=1))
    
  # print(legendplot)
  print(legendonly)
}
```
## Metadata

Loop through the specified history files and collect/organize metadata. 

```{r metadata}

extract_langtrialsexpnum <- function(filename) {
  filename_levels <- str_split_1(filename, "\\\\")
  num_levels <- length(filename_levels)
  filename_only <- filename_levels[num_levels]
  containing_folder <- filename_levels[num_levels-1]
  
  if (grepl("FullHistory", filename_only)) {
    # output is from OTSoft
    langstring <- substr(containing_folder, 1, 2)
    expnumstring <- substr(containing_folder, 3, 5)
  } else {
    # output is from my Python script
    langstring <- substr(filename_only, 1, 2)
    expnumstring <- substr(filename_only, 3, 5)
  }
  trialsstring <- switch(
    strtoi(substr(expnumstring, 3, 3)),
    "100",
    "500",
    "1000",
    "5000",
    "10000"
  )
  return(c(langstring, trialsstring, expnumstring))
}

makeplot <- function(filename) {
  info <- extract_langtrialsexpnum(filename)
  lang <- info[1]
  trials <- info[2]
  expnum <- info[3]
  df <- prepare_data(filename)
  labels <- levels(df$constraint)
  cons <- removeconvalue(labels)
  colourmap <- map_colours(lang, cons, labels)
  plot_trajectories(df, colourmap, lang, trials, expnum)
  plot_legends(df, colourmap, lang, trials, expnum)
}

historyfiles <- c(
  # "TODO",
  "zzzNOPE_(old_and-or_nonempty_is_backwards)\\20240104_Magri_outputs\\Fi153_python_OTSoft-PDDP-Fin_GLA\\Fi153_OTSoft-PDDP-Fin_GLA_HISTORY1000_forR.txt",
  "zzzNOPE_(old_and-or_nonempty_is_backwards)\\20240104_Magri_outputs\\NE153_python_OTSoft-PDDP-NEst_GLA\\NE153_OTSoft-PDDP-NEst_GLA_HISTORY1000_forR.txt",
  "zzzNOPE_(old_and-or_nonempty_is_backwards)\\20240105_GLA_KVoutputs\\NE103_FilesForOTSoft-PDDP-NEst_GLA\\OTSoft-PDDP-NEst_GLAFullHistory.xls"
)

for (hfile in historyfiles) {
  suppressWarnings(makeplot(hfile))
}



```



